---
title: "Weather Events Impact Analysis"
author: "Jason R. Battles"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This brief study involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The goal of this study is to explore the NOAA Storm Database and answer some questions about severe weather events and their impact on fatalities, injuries, and economies.

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

This report may be considered by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events.

### Prepare Environment
The analysis environment must be prepared with the appropriate libraries
```{r}
install.packages("downloader")
library(downloader)
library(ggplot2)
library(grid)
```

## Data Processing

### Data Load
Download the compressed data file. Decompress the file and assign the results to a raw data frame (dataRaw).
```{r cache=TRUE}

download(url="https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", destfile="../data/StormData.csv.bz2")

dataRaw <- read.csv(bzfile("../data/StormData.csv.bz2"))
```

### Data Verification
There should be 90,2297 observations across 37 variables
```{r}
str(dataRaw)
```

### Data Transformation
Now that the raw data is loaded into a raw data frame, subset the data to create a new data frame with a prioritiezed list of fatalities by event type.
```{r}
fatals <- aggregate(dataRaw$FATALITIES, by=list(dataRaw$EVTYPE), FUN=sum)
colnames(fatals) <- c("EventType", "Fatalities")
fatals <- fatals[with(fatals, order(-Fatalities)),]
mostfatal <- fatals[1:12,]

## Change the order of the factor levels by ordering them explicitly.  Allows ascending ggplot pgraph order. 
mostfatalordered <- mostfatal
mostfatalordered$EventType <- factor(mostfatalordered$EventType, levels = mostfatalordered$EventType[order(mostfatalordered$Fatalities)])
```

Now subset the data again to create a new data frame with a prioritized list of injuries by event type.
```{r}
injuries <- aggregate(dataRaw$INJURIES, by=list(dataRaw$EVTYPE), FUN=sum)
colnames(injuries) <- c("EventType", "Injuries")
injuries <- injuries[with(injuries, order(-Injuries)),]
mostinjuries <- injuries[1:12,]

## Change the order of the factor levels by ordering them explicitly.  Allows ascending ggplot graph order. 
mostinjordered <- mostinjuries
mostinjordered$EventType <- factor(mostinjordered$EventType, levels = mostinjordered$EventType[order(mostinjordered$Injuries)])
```

Add more readable column headers, transform the observations to understandable numbers. Subset the data again to create new data frames with a prioritized list of property damage and crop damage by event type.  Economic values, some of which are in Billions (B) and Millions (M), and Thousands (K), are all converted to Millions

```{r}
damage <- dataRaw[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

## Convert damage values to Millions
damage$PropDamage <- damage$PROPDMG
hasB_prop <- which(damage$PROPDMG > 0 & damage$PROPDMGEXP == "B")
for (i in hasB_prop) damage[i, "PropDamage"] <- damage[i, "PROPDMG"] * 1000

hasK_prop <- which(damage$PROPDMG > 0 & damage$PROPDMGEXP == "K")
for (i in hasK_prop) damage[i, "PropDamage"] <- damage[i, "PROPDMG"] / 1000

damage$CropDamage <- damage$CROPDMG
hasB_crop <- which(damage$CROPDMG > 0 & damage$CROPDMGEXP == "B")
for (i in hasB_crop) damage[i, "CropDamage"] <- damage[i, "CROPDMG"] * 1000

hasK_crop <- which(damage$CROPDMG > 0 & damage$CROPDMGEXP == "K")
for (i in hasK_crop) damage[i, "CropDamage"] <- damage[i, "CROPDMG"] / 1000

propdamage <- aggregate(damage$PropDamage, by=list(damage$EVTYPE), FUN=sum)
cropdamage <- aggregate(damage$CropDamage, by=list(damage$EVTYPE), FUN=sum)
colnames(propdamage) <- c("EventType", "Damage")
colnames(cropdamage) <- c("EventType", "Damage")
propdamage <- propdamage[with(propdamage, order(-Damage)),]
cropdamage <- cropdamage[with(cropdamage, order(-Damage)),]

combined <- merge(propdamage, cropdamage, by="EventType", all.x=TRUE, all.y=TRUE)
combined <- transform(combined, TotalDamage = Damage.x + Damage.y)
colnames(combined) <- c("EventType", "Property", "Crop", "Total")
combined <- combined[with(combined, order(-Total, -Property, -Crop)),]

## Identify the Top 12 damaging events -- Total Damage, Property Damage, Crop Damage
mostdamage <- combined[1:12,]
mostpropdamage <- propdamage[1:12,]
mostcropdamage <- cropdamage[1:12,]

## Change the order of the factor levels by ordering them explicitly.  Allows ascending ggplot pgraph order. 
mostdamageOrd <- mostdamage
mostdamageOrd$EventType <- factor(mostdamageOrd$EventType, levels = mostdamageOrd$EventType[order(mostdamageOrd$Total)])

## Change the order of the factor levels by ordering them explicitly.  Allows ascending ggplot pgraph order. 
mostpropdamageOrd <- mostpropdamage
mostpropdamageOrd$EventType <- factor(mostpropdamageOrd$EventType, levels = mostpropdamageOrd$EventType[order(mostpropdamageOrd$Damage)])

## Change the order of the factor levels by ordering them explicitly.  Allows ascending ggplot pgraph order. 
mostcropdamageOrd <- mostcropdamage
mostcropdamageOrd$EventType <- factor(mostcropdamageOrd$EventType, levels = mostcropdamageOrd$EventType[order(mostcropdamageOrd$Damage)])

```


## Results

#### 1. Across the United States, which types of events are most harmful with respect to population health?

By a very wide margin, *Tornadoes* cause the most population fatalities and injuries for all weather events (*Graph #1*). In regards to Fatalaties, Excessive Heat was #2 and Flash Floods was #3.  

In regards to Injuries, after *Tornadoes*, Marine Straight-Line Winds (TSTM) was #2, and Floods was #3 (*Graph #2*).

```{r}
# function for displaying any plot in the grid
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)

plot1 <- ggplot(mostfatalordered, aes(x = EventType, y = Fatalities)) + geom_bar(stat= "identity", fill = "red", col = "black") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + labs(title = "Total Fatalities by Event Type") + xlab("")
plot2 <- ggplot(mostinjordered, aes(x = EventType, y = Injuries)) + geom_bar(stat= "identity", fill = "orange", col = "black") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + labs(title = "Total Injuries by Event Type") + xlab("")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(plot1, vp = vplayout(1, 1))
print(plot2, vp = vplayout(1, 2))
```



#### 2. Across the United States, which types of events have the greatest economic consequences?
*Floods* caused the most Total Damage (*Graph #3*).  *Floods* also cause the most in Total Property Damage (*Graph #4*).  Howoever, *Droughts* cause the Most Crop Damage (*Graph #3*)
```{r}
plot3 <- ggplot(mostdamageOrd, aes(x=EventType, y=Total)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Damage by Event Type") + xlab("")
plot4 <- ggplot(mostpropdamageOrd, aes(x=EventType, y=Damage)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Property Damage by Event Type") + xlab("")
plot5 <- ggplot(mostcropdamageOrd, aes(x=EventType, y=Damage)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Crop Damage by Event Type") + xlab("")

grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(plot3, vp = vplayout(1, 1))
print(plot4, vp = vplayout(2, 1))
print(plot5, vp = vplayout(3, 1))
```


