---
title: "Weather Events Impact Analysis"
author: "Jason R. Battles"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This brief study involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The basic goal of this study is to explore the NOAA Storm Database and answer some basic questions about severe weather events and their impact on fatalities, injuries, and economic impact.

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

This report may be considered by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events.

## Data Processing

### Prepare Environment

```{r}
install.packages("downloader")
library(downloader)
library(ggplot2)
library(grid)
```

### Data Load
Download the compressed data file. Decompress the file and assign the results to a data frame.
```{r cache=TRUE}

download(url="https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", destfile="../data/StormData.csv.bz2")

dataRaw <- read.csv(bzfile("../data/StormData.csv.bz2"))
```

### Data Verification
There should be 90,2297 observations of 37 variables
```{r}
str(dataRaw)
```

### Data Transformation
Now that the raw data is loaded into a raw data frame, subset the data to create a new data frame with a prioritiezed list of fatalities by event type.
```{r}
fatals <- aggregate(dataRaw$FATALITIES, by=list(dataRaw$EVTYPE), FUN=sum)
colnames(fatals) <- c("EventType", "Fatalities")
fatals <- fatals[with(fatals, order(-Fatalities)),]
mostfatal <- fatals[1:12,]
mostfatalordered <- mostfatal
mostfatalordered$EventType <- factor(mostfatalordered$EventType, levels = mostfatalordered$EventType[order(mostfatalordered$Fatalities)])
```
Now subset the data again to create a new data frame with a prioritiezed list of injuries by event type.
```{r}
injuries <- aggregate(dataRaw$INJURIES, by=list(dataRaw$EVTYPE), FUN=sum)
colnames(injuries) <- c("EventType", "Injuries")
injuries <- injuries[with(injuries, order(-Injuries)),]
mostinjuries <- injuries[1:12,]
mostinjordered <- mostinjuries
mostinjordered$EventType <- factor(mostinjordered$EventType, levels = mostinjordered$EventType[order(mostinjordered$Injuries)])
```

Add more readable column headers, transform the observations to understandable numbers. Subset the data again to create new data frames with a prioritized list of property damage and crop damage by event type.  Economic values, some of which are in Billions (B) and Millions (M), are coverted to Thousands (K)

```{r}
damage <- dataRaw[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

damage$PropDamage <- damage$PROPDMG
hasB_prop <- which(damage$PROPDMG > 0 & damage$PROPDMGEXP == "B")
for (i in hasB_prop) damage[i, "PropDamage"] <- damage[i, "PROPDMG"] * 1000

hasK_prop <- which(damage$PROPDMG > 0 & damage$PROPDMGEXP == "K")
for (i in hasK_prop) damage[i, "PropDamage"] <- damage[i, "PROPDMG"] / 1000

damage$CropDamage <- damage$CROPDMG
hasB_crop <- which(damage$CROPDMG > 0 & damage$CROPDMGEXP == "B")
for (i in hasB_crop) damage[i, "CropDamage"] <- damage[i, "CROPDMG"] * 1000

hasK_crop <- which(damage$CROPDMG > 0 & damage$CROPDMGEXP == "K")
for (i in hasK_crop) damage[i, "CropDamage"] <- damage[i, "CROPDMG"] / 1000

propdamage <- aggregate(damage$PropDamage, by=list(damage$EVTYPE), FUN=sum)
cropdamage <- aggregate(damage$CropDamage, by=list(damage$EVTYPE), FUN=sum)
colnames(propdamage) <- c("EventType", "Damage")
colnames(cropdamage) <- c("EventType", "Damage")
propdamage <- propdamage[with(propdamage, order(-Damage)),]
cropdamage <- cropdamage[with(cropdamage, order(-Damage)),]

combined <- merge(propdamage, cropdamage, by="EventType", all.x=TRUE, all.y=TRUE)
combined <- transform(combined, TotalDamage = Damage.x + Damage.y)
colnames(combined) <- c("EventType", "Property", "Crop", "Total")
combined <- combined[with(combined, order(-Total, -Property, -Crop)),]
mostdamage <- combined[1:12,]

mostpropdamage <- propdamage[1:12,]
mostcropdamage <- cropdamage[1:12,]

mostdamageOrd <- mostdamage
mostdamageOrd$EventType <- factor(mostdamageOrd$EventType, levels = mostdamageOrd$EventType[order(mostdamageOrd$Total)])

mostpropdamageOrd <- mostpropdamage
mostpropdamageOrd$EventType <- factor(mostpropdamageOrd$EventType, levels = mostpropdamageOrd$EventType[order(mostpropdamageOrd$Damage)])

mostcropdamageOrd <- mostcropdamage
mostcropdamageOrd$EventType <- factor(mostcropdamageOrd$EventType, levels = mostcropdamageOrd$EventType[order(mostcropdamageOrd$Damage)])

```


## Results

#### 1. Across the United States, which types of events are most harmful with respect to population health?

By a very wide margin, tornadoes cause the most population fatalities and injuries for all weather events. In regards to Fatalaties, Excessive Heat was #2 and Flash Floods was #3.  In regards to Injuries, Marine Straight-Line Winds (TSTM) was #2, and Floods was #3.

```{r}
# function for displaying any plot in the grid
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)

plot1 <- ggplot(mostfatalordered, aes(x = EventType, y = Fatalities)) + geom_bar(stat= "identity", fill = "red", col = "black") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + labs(title = "Total Fatalities by Event Type") + xlab("")
plot2 <- ggplot(mostinjordered, aes(x = EventType, y = Injuries)) + geom_bar(stat= "identity", fill = "orange", col = "black") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + labs(title = "Total Injuries by Event Type") + xlab("")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(plot1, vp = vplayout(1, 1))
print(plot2, vp = vplayout(1, 2))
```



#### 2. Across the United States, which types of events have the greatest economic consequences?
Floods caused the most Total Damage (graph #1).  Floods also cause the most in Total Property Damage (graph #2).  Howoever, droughts cause the Most Crop Damage (graph #3)
```{r}
plot3 <- ggplot(mostdamageOrd, aes(x=EventType, y=Total)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Damage by Event Type") + xlab("")
plot4 <- ggplot(mostpropdamageOrd, aes(x=EventType, y=Damage)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Property Damage by Event Type") + xlab("")
plot5 <- ggplot(mostcropdamageOrd, aes(x=EventType, y=Damage)) + geom_bar(stat= "identity") + theme(axis.text.x = element_text(size=8, angle=45, hjust=1)) + ylab("$MM") + labs(title = "Total Crop Damage by Event Type") + xlab("")

grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(plot3, vp = vplayout(1, 1))
print(plot4, vp = vplayout(2, 1))
print(plot5, vp = vplayout(3, 1))
```


